\plaintitle

### 2017-01-04 Paper/proceedings.tex
JH edits
Made changes based on JH comments, need to fix some bad widows and
orphans for paragraphs on pgs 7 and 8

Similarly, visual design choices (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}. The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}, as wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts. Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.

@@

We investigated three types of common bivariate visualizations: scatter plots, line graphs, and area charts (see Fig. \ref{fig:types}). In each experiment, we presented participants with a visualization and asked them to interactively adjust trend lines to best fit the data.  Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}. Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed). We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. While other studies investigate the use of more expressive designs (such as free-form drawing)~\cite{kim2017}, we sought to disambiguate the model selection task (e.g., the selection of a a linear or non-linear model) from the model fitting task (adjusting the parameters of the selected model). Our design also reduces impact of factors such as noise from motor movements or heterogeneous input devices.

@@

For each point set, we placed 100 points along a particular trend, regularly spaced in x. This both affords the use of line graphs and area charts for displaying this data (as they require that points be one to one), and aligns well with time series data, a common domain for estimating (temporal) trends and relationships. For each point set, we created a set of residual values, sampled evenly from a Gaussian. The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig. \ref{fig:sigmas}). We permuted this set of residuals, and then applied them to the original points. As heteroskedasticity introduced through permutation could alter the trend away from the target trend, we performed rejection sampling to ensure that the slope of the trend of the resulting points was within $10^{-7}$ of the target. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Fig. \ref{fig:trendtypes} shows these different trend types on the same set of residuals.

@@

Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}. Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise interactions. The interquartile mean of absolute error of was $0.06$ for linear and quadratic fits, and $0.07$ for trigonometric fits (see Fig. \ref{fig:exp1types}). This result suggests that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.

@@

Prior work in graphical perception often measures performance as the absolute log error: e.g., Cleveland \& McGill~\cite{cleveland1984graphical} calculate their performance metric as $\log_2 (|\text{error}| + \frac{1}{8}$). As a point of comparison, the interquartile mean of the absolute log error across all conditions in this experiment was $2.4$. Differences in methodology and measurement discourage statistical inferences comparing this value with those in similar experiments. Nevertheless, error rates for regression by eye are at least in the same magnitude as those observed in other graphical perception tasks in both lab and crowdsourced studies. For instance, Heer \& Bostock~\cite{heer2010crowdsourcing} report an absolute log error of $2.5$ for proportional judgments in treemaps, and Cleveland \& McGill~\cite{cleveland1984graphical} report an absolute log error of $2.4$ for proportional judgments in stacked bars (although compare to a log error of estimation of relative lengths of lines of $1.1$). This similarity suggests that, despite requiring estimation of aggregate statistical information, regression by eye results in judgments that are accurate enough for many practical purposes, comparable with the expected accuracy in comparing individual values in visualizations.

@@

Our results also fail to support our third hypothesis: \textbf{There was no significant impact of outlier location on performance}. There is a confound between outlier location and OLS line: outliers on the ends of the data ranges create larger shifts in slope than outliers in the center, which may only create shifts in intercept. To evaluate our third hypothesis, we therefore defined a parameterized performance function $\frac{\text{Estimated Slope}-\text{Robust Slope}}{\text{OLS Slope} - \text{Robust Slope}}$ that contextualizes a participant estimate as an interpolation between the robust slope (value of $0$) and the OLS slope (value of $1$). A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant ID as a random factor, found no significant effect of outlier location on interpolative value ($F(2,3441)=0.63$, $p=0.53$).

@@

	\item \emph{Designers should avoid area charts for data where the perception of trends is important.} The visual asymmetry caused by filling in the area under the line results in a corresponding asymmetry in judgment of trend: an underestimation in perceived trend. This within-the-area bias is undesirable in many cases, but can be countered by visual encodings (such as scatter plots and line charts) that have no such visual asymmetry.
\end{enumerate}

@@

Our future work is focused in three areas. First, we wish to examine the impact of annotations relevant to regression (such as confidence bands, error bars, and curve boxplots~\cite{mirzargar2014curve}) on regression by eye. Can sufficient information promote caution in judgments of trend? Second, we wish to examine the impact of different rhetorical framings and presentations on regression by eye. Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings could similarly impact statistical judgments. For instance, outliers from a source that is perceived as credible could be weighted higher than outliers with more suspect provenance. Likewise, preconceptions about the volatility of data from certain domains may color how experts estimate trends. Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities. Cognitive and perceptual biases are difficult to overcome, and may require exploration of novel forms of visualization~\cite{micallef2012assessing}.

@@

This work supported by a Moore Foundation Data-Driven Investigator Award.

>>>>

Similarly, visual design choices (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}. The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}: narrow aspect ratios can result in overestimation in severity of trends, and wider aspect ratios in underestimation of severity. Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.

@@

We investigated three types of common bivariate visualizations: scatter plots, line graphs, and area charts (see Fig. \ref{fig:types}). In each experiment, we presented participants with a visualization and asked them to interactively adjust trend lines to best fit the data.  Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}. Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed). We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. While other studies investigate the use of more expressive designs (such as free-form drawing)~\cite{kim2017}, we sought to disambiguate the model selection task (e.g., the selection of a a linear or non-linear model) from the model fitting task (adjusting the parameters of the selected model). Our design also reduces the impact of factors such as noise from motor movements or heterogeneous input devices.

@@

For each point set, we placed 100 points along a particular trend, regularly spaced in x. This both affords the use of line graphs and area charts for displaying this data (as they require that points be one to one), and aligns well with time series data, a common domain for estimating (temporal) trends and relationships. For each point set, we created a set of residual values, sampled evenly from a Gaussian. The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig. \ref{fig:sigmas}). We permuted this set of residuals, and then applied them to the original points. As heteroskedasticity introduced through permutation could alter the trend away from the target trend, we performed rejection sampling to ensure that the slope of the trend of the resulting points was within $10^{-7}$ of the target. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Fig. \ref{fig:trendtypes} shows these different trend types.

@@

Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}. Fit type was not a significant main effect ($F(2,4554)=2.60$, $p=0.074$), and post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise interactions. The interquartile mean of absolute error of was $0.06$ for linear and quadratic fits, and $0.07$ for trigonometric fits (see Fig. \ref{fig:exp1types}). This suggests that the relative unfamiliarity of non-linear trends does not result in poorer performance at regression by eye.

@@

Prior work in graphical perception often measures performance as the absolute log error: e.g., Cleveland \& McGill~\cite{cleveland1984graphical} calculate their performance metric as $\log_2 (|\text{error}| + \frac{1}{8}$). As a point of comparison, the interquartile mean of the absolute log error across all conditions in this experiment was $2.4$. Differences in methodology and measurement discourage statistical inferences comparing this value with those in similar experiments. Nevertheless, error rates for regression by eye are at least in the same magnitude as those observed in other graphical perception tasks in both lab and crowdsourced studies. For instance, Heer \& Bostock~\cite{heer2010crowdsourcing} report an absolute log error of $2.5$ for proportional judgments in treemaps, and Cleveland \& McGill~\cite{cleveland1984graphical} report an absolute log error of $2.4$ for proportional judgments in stacked bars (although compare to a log error of estimation of relative lengths of lines of $1.1$). This similarity suggests that, despite requiring estimation of aggregate statistical information, regression by eye results in judgments that are accurate enough for many practical purposes, comparable with accuracy in comparing individual values in visualizations.

@@

Our results also fail to support our third hypothesis: \textbf{There was no significant impact of outlier location on performance}. There is a confound between outlier location and OLS line: outliers on the ends of the data ranges create larger shifts in slope than outliers in the center, which may only create shifts in intercept. To evaluate our third hypothesis, we therefore defined a parameterized performance function $\frac{\text{Estimated Slope}-\text{Robust Slope}}{\text{OLS Slope} - \text{Robust Slope}}$ that contextualizes a participant estimate as an interpolation between the robust slope (value of $0$) and the OLS slope (value of $1$). A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant id as a random factor, found no significant effect of outlier location on interpolative value ($F(2,3441)=0.63$, $p=0.53$).

@@

	\item \emph{Designers should avoid area charts when the perception of trends is important.} The visual asymmetry caused by filling in the area under the line results in a corresponding asymmetry in judgment of trend: an underestimation in perceived trend. This within-the-area bias is undesirable in many cases, but can be countered by visual encodings (such as scatter plots and line charts) that have no such visual asymmetry.
\end{enumerate}

@@

Our future work is focused in three areas. First, we wish to examine the impact of annotations relevant to regression (such as confidence bands, error bars, and curve boxplots~\cite{mirzargar2014curve}) on regression by eye. Can sufficient information promote caution in judgments of trend? Second, we wish to examine the impact of different rhetorical framings and presentations on regression by eye. Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings could similarly impact statistical judgments. For instance, outliers from a credible source could be weighted higher than outliers with more suspect provenance. Likewise, preconceptions about the volatility of data from certain domains may color how experts estimate trends. Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities. Cognitive and perceptual biases are difficult to overcome, and may require exploration of new visualization designs~\cite{micallef2012assessing}.

@@

This work supported by a Moore Foundation Data-Driven Discovery Investigator Award.
